---
title: "Storm Data Analysis"
author: "Rohit Jain"
date: "8/28/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.In this data analysis we will be analysing the effect of these events on human and property with help of R. 


## Data Processing
```{r dataprocessing, cache=TRUE, warning=FALSE }
#loading the dataset
dt <- read.table("StormData.csv.bz2", sep = ",", header = TRUE)

#preprocessing the Event type:
dt$EVTYPE <- tolower(dt$EVTYPE)
dt$EVTYPE[grepl(pattern = "*tstm*", dt$EVTYPE)] <- "Tstm Wind"
dt$EVTYPE[grepl(pattern = "^tropical strom*", dt$EVTYPE)] <- "Tropical Storm"
dt$EVTYPE[grepl(pattern = "*flood*", dt$EVTYPE)] <- "Flood"
dt$EVTYPE[grepl(pattern = "*hail*", dt$EVTYPE)] <- "Hail"
dt$EVTYPE[grepl(pattern = "*rain*", dt$EVTYPE)] <- "Rain"
dt$EVTYPE[grepl(pattern = "*thunder*", dt$EVTYPE)] <- "Thunderstorm"
dt$EVTYPE[grepl(pattern = "*snow*", dt$EVTYPE)] <- "Snow Related"
dt$EVTYPE[grepl(pattern = "*surf*", dt$EVTYPE)] <- "Surf Related"
dt$EVTYPE[grepl(pattern = "*microburst*", dt$EVTYPE)] <- "Dry Microburst"
dt$EVTYPE[grepl(pattern = "*funnel*", dt$EVTYPE)] <- "Funnel Cloud"
dt$EVTYPE[grepl(pattern = "*glaze*", dt$EVTYPE)] <- "Glaze"
dt$EVTYPE[grepl(pattern = "*shower*", dt$EVTYPE)] <- "Rain"
dt$EVTYPE[grepl(pattern = "*high wind*", dt$EVTYPE)] <- "High Winds"
dt$EVTYPE[grepl(pattern = "*hurricane*", dt$EVTYPE)] <- "Hurricane"
dt$EVTYPE[grepl(pattern = "*light*", dt$EVTYPE)] <- "Lighting"
dt$EVTYPE[grepl(pattern = "*summary*", dt$EVTYPE)] <- "Summary related Entry"
dt$EVTYPE[grepl(pattern = "*tornado*", dt$EVTYPE)] <- "Tornado"
dt$EVTYPE[grepl(pattern = "*waterspout*", dt$EVTYPE)] <- "Waterspout"
dt$EVTYPE[grepl(pattern = "*wild*", dt$EVTYPE)] <- "Wild Fire"
dt$EVTYPE[grepl(pattern = "*record*", dt$EVTYPE)] <- "Record Conditions"
dt$EVTYPE[grepl(pattern = "*blizzard*", dt$EVTYPE)] <- "Blizzard Conditions"
dt$EVTYPE[grepl(pattern = "*cold*", dt$EVTYPE)] <- "Cold Related"
dt$EVTYPE[grepl(pattern = "*extreme*", dt$EVTYPE)] <- "Extreme Conditions"
dt$EVTYPE[grepl(pattern = "*frost*", dt$EVTYPE)] <- "Frost"
dt$EVTYPE[grepl(pattern = "*gusty*", dt$EVTYPE)] <- "Gusty Wind"
dt$EVTYPE[grepl(pattern = "^heat*", dt$EVTYPE)] <- "Heat Related"

# Pre processing for economic effect:

dt$PROPDMGEXP <- as.character(dt$PROPDMGEXP)
dt$PROPDMGEXP[dt$PROPDMGEXP %in% c("m","M")] <- 6
dt$PROPDMGEXP[dt$PROPDMGEXP %in% c("k","K")] <- 3
dt$PROPDMGEXP[dt$PROPDMGEXP %in% c("h","H")] <- 2
dt$PROPDMGEXP[dt$PROPDMGEXP %in% c("b","B")] <- 9
dt$PROPDMGEXP[dt$PROPDMGEXP %in% c('-','?','+',' ')] <- 0
dt$PROPDMGEXP <- as.numeric(dt$PROPDMGEXP)
dt$propdmg <- dt$PROPDMG * 10^dt$PROPDMGEXP

dt$CROPDMGEXP <- as.character(dt$CROPDMGEXP)
dt$CROPDMGEXP[dt$CROPDMGEXP %in% c("m","M")] <- 6
dt$CROPDMGEXP[dt$CROPDMGEXP %in% c("k","K")] <- 3
dt$CROPDMGEXP[dt$CROPDMGEXP %in% c("h","H")] <- 2
dt$CROPDMGEXP[dt$CROPDMGEXP %in% c("b","B")] <- 9
dt$CROPDMGEXP[dt$CROPDMGEXP %in% c('-','?','+',' ')] <- 0
dt$CROPDMGEXP <- as.numeric(dt$CROPDMGEXP)
dt$cropdmg <- dt$CROPDMG * 10^dt$CROPDMGEXP
```

## Data Analysis

### Effect of the Events on Humans(Fatalities and Inturies)
In this part of the document we will be extracting out the top 10 event for the Dataset, which has the worst effect on human in context of fatality and Injury.

```{r HumanEffect, warning=FALSE}
library(plyr)
humanEffect <- ddply(dt, .(EVTYPE), summarize,
                     fatalities = sum(FATALITIES),
                     injuries= sum(INJURIES))
#Event type
fatal_events <- head(humanEffect[order(humanEffect$fatalities, decreasing = T), ], 10)
injury_events <- head(humanEffect[order(humanEffect$injuries, decreasing = T), ], 10)
```

#### Top 10 Events having worst effect on human health(Injuries and Fatalities)

``` {r listhumEffect, warning=FALSE}
fatal_events[, c("EVTYPE", "fatalities")]
injury_events[, c("EVTYPE", "injuries")]
```

### Effect of the Events on Property and Crops
In this part of the document we will be extracting out the top 10 event for the Dataset, which has the worst economic effect on Property and Crops.

``` {r economicEffect, warning=FALSE}

ecoEffect <- ddply(dt, .(EVTYPE), summarize,
                   property = sum(propdmg),
                   crops= sum(cropdmg))

prop_dmg <- head(ecoEffect[order(ecoEffect$property, decreasing = TRUE),],10)
prop_dmg$property[is.na(prop_dmg$property)] <- 0
crop_dmg <- head(ecoEffect[order(ecoEffect$crops, decreasing = TRUE ),],10)
crop_dmg$crops[is.na(crop_dmg$crops)] <- 0
```

#### Top 10 Events having worst economic effect

```{r listecoEffect, warning=FALSE}
prop_dmg[, c("EVTYPE", "property")]
crop_dmg[, c("EVTYPE", "crops")]
```

## Result

In this section we will be depicting the outcome of the analysis in the form graphs where the "Event Vs Effect" relationship is shown.

### Human Effect

``` {r humanEffectgraph , warning = FALSE}

library(ggplot2)
library(gridExtra)

fatalPlot <- ggplot(data = fatal_events, aes(EVTYPE, fatalities)) + 
              geom_bar(stat = "identity", fill = "red") +
              labs(title = "Event Vs Effect", x = "Event" , y = "Fatalities") +
              theme(axis.text.x = element_text(angle = 90, face = "bold", colour = "black"))

injuryPlot <- ggplot(data = injury_events, aes(EVTYPE, injuries)) +
              geom_bar(stat = "identity", fill = "blue") +
              labs(title = "Event Vs Effect", x = "Event" , y = "Injuries") +
              theme(axis.text.x = element_text(angle = 90, face = "bold", colour = "black"))

grid.arrange(fatalPlot, injuryPlot, nrow = 1)
```

### Economic Effect

```{r ecoEffectgraph, warning=FALSE}
library(ggplot2)
library(gridExtra)

propPlot <- ggplot(data = prop_dmg, aes(EVTYPE, log(property))) + 
  geom_bar(stat = "identity", fill = "red") +
  labs(title = "Event Vs Effect", x = "Event" , y = "Property damage") +
  theme(axis.text.x = element_text(angle = 90, face = "bold", colour = "black"))

cropPlot <- ggplot(data = crop_dmg, aes(EVTYPE, log(crops))) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Event Vs Effect", x = "Event" , y = "Crop damage") +
  theme(axis.text.x = element_text(angle = 90, face = "bold", colour = "black"))

grid.arrange(propPlot, cropPlot, nrow = 1)
```
